Staged Abstract Interpreters

### 2019-02-19 paper/10_intro.tex
intro

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask --- starting from an interpreter, \textbf{can we derive sound abstract
interpreters but it works like a compiler and generates fast analysis code?}
Similar to the concrete interpreter, for an abstract interpreter
$\widehat{\texttt{eval}}$ that takes a program, an abstract environment and
returns abstract values, shown as the following:

@@

We would like to have a specialized abstract interpreter
$\widehat{\texttt{eval}}_{\texttt{e0}}$ given a program $e_0$ statically, but
leave the abstract environment $\rho$ dynamic.

@@

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static analyzer
without any intrusive changes and soundness compromises?} Static analysis is
known as a trade-off game between precision and efficiency, we argue that
applying meta-programming (particularly, multi-stage programming) with abstract
interpreters is an effective approach to improve the latter one while fix the
former one. Specialization removes the overhead of abstractions introduced by
high-level programming, but not changes the meaning of the program -- in this
case, the abstract semantics implemented by the analyzer.
Indeed, specializing analysis with respect to a program by partial evaluation
is not new \cite{damian1999partial, amtoft1999partial,
  Boucher:1996:ACN:647473.727587, ashley:practical}, despite the fact that these
work are ad-hoc on a particular analysis or requiring significant changes to
the analysis. For example, abstract compilation \cite{Boucher:1996:ACN:647473.727587}
requires the whole analyzer to be rewritten as closure generation from. In this
paper, we show a systematic approach to meta-programming achieves the
performance goal with less intrusive changes; in fact, a case study
(Section~\ref{cs_ac}) that compared with abstract compilation shows, by
utilizing type-directed binding-time annotations, the analyzer program does not
need any changes.

\todo{modular analysis}

@@

\subsection{Contributions}

This paper makes the following contributions:

\begin{itemize}
  \item Starting from a generic definition of an interpreter for a higher-order
    functional language, whose syntax and semantics are described in
    Section~\ref{bg_lang}, we show its concrete interpreter
    (Section~\ref{unstaged_conc}), staged concrete interpreter
    (Section~\ref{stagedinterp}), abstract interpreter (Section~\ref{unstaged_abs})
    and staged abstract interpreter (Section~\ref{sai}). Our design allows to
    derive these four different kinds of interpreters by mechanically changing the
    underlying semantics or stage annotations.
    %We later extend the staged abstract interpreter to support imperative
    %language features (Section~\ref{cases_imp}), such as states and loops.

  \item As shown in Section~\ref{sai}, we demonstrate that if we apply the
    abstract interpreters on open programs, it not only improves the
    \textit{efficiency} but also the \textit{scalability}, which are two major
    issues in static analysis.

  \item To demonstrate the applicability and usefulness of the proposed
    approach, we conduct three case studies in Section~\ref{cases_study}:

    \subitem (1) We revisit Abstract Compilation \cite{Boucher:1996:ACN:647473.727587} in Section~\ref{cs_ac}, 
    which was originally a closure generation technique applied on 0-CFA.
    We show that abstract compilation can be understood and easily implemented
    as an instance of our staged abstract interpreter. But unlike abstract
    compilation, our approach does not require the implementers to rewrite the
    analysis.

    \subitem (2) We demonstrate pushdown control-flow analysis techniques
    (Section~\ref{cfa}), such as context-sensitive analysis and more precise
    stores can be integrated to staged abstract interpreters without losing any
    precision.

    \subitem (3) We show that in an imperative language
    (Section~\ref{cases_imp}), by using an interval domain as abstract domain
    (as well as staged), the staged abstract interpreter performs numerical 
    analysis.

    % (Optional: taint/information flow analysis?)
  \item We empirically evaluate the performance improvement by staging a
    big-step abstract interpreter for control-flow analysis
    (Section~\ref{evaluation}). We compare our staged abstract interpreter with the
    unstaged version.
\end{itemize}

>>>>

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask --- starting from an interpreter, \textbf{can we derive a sound
abstract interpreter but it works like a compiler and generates fast analysis
code?} Similar to the concrete interpreter, for an abstract interpreter
$\widehat{\texttt{eval}}$ that takes a program, an abstract environment and
returns abstract values, shown as the following:

@@

Given a program $e_0$ statically, We would like to have a specialized abstract
interpreter $\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, but leave the abstract environment $\rho$ dynamic.

@@

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static
analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between precision and efficiency, we argue
that applying meta-programming (particularly, multi-stage programming) with
abstract interpreters is an effective approach to improve the latter one while
fix the former one. Specialization removes the overhead of abstractions
introduced by high-level programming, but not changes the meaning of the program
-- in this case, the abstract semantics implemented by the analyzer. Indeed,
specializing analysis with respect to a program by partial evaluation
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new idea, despite the fact that most of the previous
work are ad-hoc on a particular analysis or requiring significant changes to the
analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

\todo{modular analysis: a whole-program abstract
  interpreter can be run on different components of program separately, we
  achieve this by meta-programming mechnically}

@@

\paragraph{Contributions and Outline}

\begin{itemize}
  \item We present a unified implementation framework that naturally extends the
    first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a tiny functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation in Section~\ref{unstaged_conc}. Then we lift it to the staged
    version by adding stage annotations to the semantic operations
    (Section~\ref{stagedinterp}); and to the abstract interpreter by applying
    abstraction to the environment, store, and values. Finally, we show the
    combination of the two transformations, dubbed \textit{staged abstract
    interpreters}, can be easily obtained (Section~\ref{sai}).

  \item \todo{modular analysis} As shown in Section~\ref{sai}, we demonstrate
    that if we apply the abstract interpreters on open programs, it not only
    improves the \textit{efficiency} but also the \textit{scalability}, which are
    two major issues in static analysis.

  \item In Section~\ref{cases_study}, we conduct three case studies to
    demonstrate the applicability and usefulness of the proposed approach to
    specialize analyses.
    (1) We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} technique in Section~\ref{cs_ac}.
    Our approach and AC are able to achieve the same goal, however, as we show,
    one of the advantage of staging through types is it does not require
    whole-analyzer refactoring.
    (2) We extend the staged abstract interpreter shown in Section~\ref{sai} to
    two variants of control-flow analysis: context-sensitive pushdown
    control-flow analysis and \todo{precise-store?}.
    (3) In Section~\ref{cases_imp}, we plug a \note{staged numerical domain}
    \todo{widening} into the staged abstract interpreter, demonstrating that a
    numerical analysis on imperative languages can be also implemented and
    speeded-up in this manner.

  \item Based on the instantiation of control-flow analysis, we empirically
    evaluate the performance improvement by staging and specialization
    (Section~\ref{evaluation}). We compare the both context-insensitive and
    \note{context-sensitive} analysis.

\end{itemize}
